Concurrent Programming in Harmony

### 2021-12-24 book/paper.tex
Made a pass over Ch 14

\begin{problems}
\item Harmony programs can usually be easily translated into
Python by hand.  For example,
\autoref{fig:incpy} is a Python version of \autoref{fig:inc}.
\begin{enumerate}
\item Run \autoref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run \autoref{fig:incpy} 1000 times.
For example, if you are using the bash shell (in Linux or Mac OS X, say), you can do the following:
\begin{code}
\begin{verbatim}
for i in {1..1000}
do
    python Up.py
done
\end{verbatim}
\end{code}
If you're using Windows, the following batch script does the trick:
\begin{code}
\begin{verbatim}
FOR /L %%i IN (1, 1, 1000) DO python Up.py
PAUSE
\end{verbatim}
\end{code}
How many times does the assertion fail (if any)?
\end{enumerate}
\item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
incrementer thread increment <{count}> <{N}> times.  Run \autoref{fig:incmany}
10 times (using Python).
Report how many times the assertion fails and what the value of <{count}>
was for each of the failed runs.
Also experiment with lower values of <{N}>.
How large does <{N}> need to be for assertions to fail?
(Try powers of 10 for <{N}>.)
\item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
and run them through Harmony.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
(Also, you do not get to use the <{atomically}> keyword
or a \emph{lock}, yet.)
\end{problems}

@@

In Harmony, any statement can be preceded by the <{atomically}>
keyword.  It means that statement as a whole is to be executed atomically.
The <{atomically}> keyword can be used to specify the behavior of
methods such as \texttt{acquire} and \texttt{release}.  But
an actual executable program---such as the one in
\autoref{fig:cssynch}---should not use the <{atomically}> keyword
because---on a normal machine---it cannot be directly compiled into
machine code.
If we want to make the program executable on hardware,
we have to show how \texttt{Lock}, \texttt{acquire}, and \texttt{release}
are implemented, not just how they are specified.
\autoref{ch:spinlock} presents such an implementation.

@@

To see that invariant $\mathcal{I}_1$ is maintained, note that
\texttt{!}<{p}> = <{True}> upon entry of \texttt{test\_and\_set}
(because of the condition on the \textbf{while} loop that the
\texttt{test\_and\_set} method is invoked in).
There are two cases:
\begin{enumerate}
\item \texttt{!}<{s}> is <{False}> upon entry to \texttt{test\_and\_set}.
Then upon exit \texttt{!}<{p}> = <{False}> and \texttt{!}<{s}> = <{True}>,
maintaining the invariant.
\item \texttt{!}<{s}> is <{True}> upon entry to \texttt{test\_and\_set}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section
because we can assume, by the induction hypothesis, that $\textit{private}[i]$ is
<{True}> just before exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread <{i}> only proceeds to the critical
section if \textit{private}[<{i}>] is <{False}>, and (ii) no other thread modifies
\textit{private}[<{i}>].

@@

\autoref{fig:suspension} shows the lock interface using suspension.
It is implemented as follows:
\begin{itemize}
\item A lock maintains both a boolean indicating whether the
lock is currently acquired and a list of contexts of threads that want to
acquire the lock.
\item
\texttt{acquire}()
\index{acquire}%
acquires the lock if available and suspends the invoking thread if not.
In the latter case, the context of the thread is added to the end of the list of contexts.
Note that \textbf{stop} is called within an atomic statement block---this is
the only exception to such an atomic statement block running to completion.
While the
thread is running no other threads can run, but when the thread suspends itself
other threads can run.
\item
\texttt{release}()
\index{release}%
checks to see if any threads are waiting to acquire the lock.
If so, it uses the <{head}> and <{tail}>
methods from the \texttt{list} module (see \autoref{ap:module})
to resume the first thread that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first thread is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
<{choose}> to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents starvation.

@@

\section*{Exercises}
\begin{problems}
\item
Run \autoref{fig:incfixed} using (i) <{synch}> and (ii) \texttt{synchS}.
Report how many states were explored by Harmony for each module.
\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables <{x}>
(initially 0) and <{y}> (initially 100) that can be accessed through methods
\texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
should return a pair that sums to 100.  Add the necessary synchronization code.
\item \label{ex:trylock} Implement \texttt{tryAcquire}(\textit{b}) as an additional
interface for both the <{synch}> and \texttt{synchS} modules.
This interface is like \texttt{acquire}(\textit{b}) but never blocks.  It
returns <{True}> if the lock was available (and now acquired) or <{False}>
if the lock was already acquired.
Hint: you do not have to change the existing code.
\item People who use an ATM often first check their balance and then withdraw
a certain amount of money not exceeding their balance.  A negative balance
is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
one to check the balance and one to withdraw money.
Note that all operations on accounts are carefully protected by a lock
(i.e., there are no data races).
The \texttt{customer}
method models going to a particular ATM and withdrawing money not exceeding
the balance.
Running the code through Harmony reveals that there is a bug.
It is a common type of concurrency bug known as \emph{Time Of Check Time Of
Execution} (TOCTOE).
\index{Time Of Check Time Of Execution}
\index{TOCTOE}
In this case, by the time the withdraw operation is performed,
the balance can have changed.
Fix the code in \autoref{fig:atm}.  Note, you should leave the
customer code the same. You are only allowed to change the
\texttt{atm\_} methods, and you cannot use the \textbf{atomically}
keyword.
\end{problems}

@@

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queueconc}
uses the \texttt{alloc} module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using \texttt{malloc}() and \texttt{free}().
\texttt{malloc}(<{v}>) returns a new memory location initialized to <{v}>,
which should be released with \texttt{free}() when it is no longer in use.
The queue maintains a <{head}> pointer to the first element in the list
and a <{tail}> pointer to the last element in the list.
The <{head}> pointer is <{None}> if and only if the queue is empty.
(<{None}> is a special address value that is not the address of any
memory location.)

@@

\autoref{fig:qtest3} checks a mixture of
concurrent <{get}> and <{put}> operations.
It starts <{N}> <{getter}> threads and
<{N}> \texttt{putter} threads.
Thread <{main}> waits until all have finished,
and then checks to make sure that whatever is left
on the queue was not also consumed by some
<{getter}> thread.
Finally, it makes sure that the queue is empty
once all the values that were added
to it by the \texttt{putter} threads have been consumed.

A common mistake that people make is to use
a globally shared variable in the implementation of
a concurrent data structure.
The <{get()}> method of the queue implementation
uses a local variable <{node}>, declared using
a \textbf{let} statement, but it would be easy to
mistakenly use a global variable <{node}> instead.
If there is only one queue, this is not an issue.
So, it is important to test not just one queue at a
time, but also multiple queues.
\autoref{fig:qtest4} gives an example that
tests the <{get}> method for use of global
variables.  Even if the assertion does not fail,
Harmony might find a data race if there are global
variables shared between different queues.

@@

So, you wrote a Harmony program and Harmony reports a bug.
Often you may just be able to figure it out by staring at the code and
going through some easy scenarios, but what if you don't?
The output of Harmony can be helpful in that case.

@@

Let us run the code through the test programs in the last
chapter.  Harmony does not detect any issues with the sequential test
in \autoref{fig:qtestseq}.
(Run this using the \texttt{-m} flag like this:
\texttt{harmony -m queue=queuebroken code/qtestseq.hny})
However, when we run
the new queue code through the test in \autoref{fig:qtestpar}, Harmony
finds a problem (even without specifying a behavior).  The command line
for this is \texttt{harmony -m queue=queuebroken code/qtestpar.hny})

Before we go look at the details of what went wrong, we want to make
sure we generate the simplest scenario.  So, first we want to explore
what the smallest <{NOPS}> (number of operations or number of
threads) that causes the bug to occur.  With some experimentation,
we find that <{NOPS}> = 2 does not find a problem, but
<{NOPS}> = 3 does
(\texttt{harmony -m queue=queuebroken -c NOPS=3 code/qtestpar.hny})).
\autoref{fig:queuebroken1} shows the Harmony output.

There is quite a bit of information in the Harmony output, and while
it may seem intimidating, we have to learn to navigate through it
step-by-step.
Let's start with looking at the red text.  Harmony found a safety
violation (something bad happened during one of the possible executions),
and in particular \texttt{thread}(2) (thread T2) was trying to dereference
\texttt{?alloc["pool"][0]["lock"]}
(aka \texttt{?alloc}.\textit{pool}[0].\texttt{lock}).

The \texttt{alloc} module maintains a shared array \textit{pool} that
it uses for dynamic allocation.  Apparently T2 tried to access
\textit{pool}[0], but it does not exist, meaning that either it was not
yet allocated, or it had been freed since it was allocated.
When we look at the top half of the figure, we see that in fact thread
T1 allocated \textit{pool}[0] in turn 2, but T3 freed it in turn 4.
Looking back down, we see that T1 executed \texttt{thread}(1) and has
since terminated, while T3 is executing \texttt{thread}(3).

Looking further at the stack traces, we can see that
T3 was in the process of executing
\texttt{release}(?<{q}>.\texttt{lock}) within
<{get}>(?<{q}>).
% Looking in the top green row, we see that the value of <{q}> is
% \{ \texttt{.lock}: <{True}>, \texttt{.next}: <{None}>, \texttt{.value}: <{None}> \}.
T1 is currently executing
\texttt{acquire}(\texttt{?alloc}.\textit{pool}[0].\texttt{lock})
within <{put}>(?<{q}>, 2), but 
\texttt{alloc}.\textit{pool}[0] does not exist.
The corresponding line of Harmony code is
\textbf{atomically when not} !\textit{binsema}
in line 25 of the \texttt{sync} module.

So, how did we get there?  In the top we can see that the order of
events was the following:
\begin{enumerate}
\item[0:] initialization completed, with <{q}> being
\{ \texttt{.lock}: <{False}>, \texttt{.next}: <{None}>, \texttt{.value}: <{None}> \};
\item[1:] thread T1 (\texttt{thread}(1)) ran and finished executing
<{put}>(1) (see the output column for that clue: the thread printed that).
We can see that <{q}>.<{next}> now points to \texttt{alloc}.\textit{pool}[0],
which the thread must have allocated.  The contents is 
\{ \texttt{.lock}: <{False}>, \texttt{.next}: <{None}>, \texttt{.value}: 1 \}, as expected;
\item[2:] thread T2 (\texttt{thread}(1)) started running, calling
<{put}>(?<{q}>, 2).  We can see it got as far as putting 2 on the queue,
but it is not yet done.  It is currently trying to acquire
\texttt{alloc}.\textit{pool}[0].\texttt{lock};
\item[3:] thread T3 (\texttt{thread}(1)) started running, calling
<{get}>($?q$).  We can also see that it freed \textit{pool}[0],
and is now releasing <{q}>.\texttt{lock};
\item[4:] thread T2 resumes and tries to access \textit{pool}[0], which
no longer exists (because T3 just freed it).
\end{enumerate}

Clearly there was a race in which T2 was trying to lock
?\textit{pool}[0].\texttt{lock} (which contained the node with the value 1)
while T3 was freeing that very same node, and T2 lost the race.
More precisely, T2 was executing <{put}>(?<{q}>, 2), when T3
preempted it with <{get}>(?<{q}>) and removed the node that T2
was trying to access.  But why did the locks not prevent this?

\index{reserve debugging}%
It is time to start stepping through the code that has been executed
before this happened.  This is sometimes known as
\emph{reverse debugging}.  In fact, Harmony allows you to step through
code forwards and backwards.  In this case, we first want to see what
T2 is doing.  You can click on its first (top-left) orange box to
time-travel to that part in the execution.  Now by hitting
$<$return$>$ repeatedly, we can quickly skip through the code.
T2 first calls <{put}>(?<{q}>, 1) and then allocates a new
node initialized with a lock.  Keep stepping until it executes
$\mathit{nq} = q$.  Hit $<$return$>$ once more and inspect the
state of T2 in the lower-right corner.  You can see that variable
<{nq}> is initialized to~?<{q}>.  T2 then enters into the
\textbf{while} loop and tries to acquire
<{nq->lock}>.  This succeeds, and
next T2 executes \textbf{let} <{n = nq->next}>.
Now $n =\,?\mathtt{alloc}.\mathit{alloc}[0]$, which is not <{None}>.
It then releases <{nq->lock}>
(<{nq}> points to <{q}>).  It then sets <{nq}> to $n$, which
is still $?\mathtt{alloc}.\mathit{alloc}[0]$.  Finally, it calls
<{acquire(?nq->lock)}>.
But before it can complete that operation, T3 runs next.

T3 chooses "get" and then goes on to invoke $\mathtt{get}(?q)$.
This first successfully acquires <{q}->lock}>.
T3 then finds out that <{q->next}> points to <{alloc.pool[0]}>.
T3 sets <{node}> to <{alloc.pool[0]}> as well and sets
<{q->{next}> to <{node->next}>.
T3 sets <{result}> to <{node->value}> (which is 1)
and then frees <{node}>.  This is where the problem is---T2 is
about to acquire the lock in that same node.

To fix the code without changing the data structure, we can use
hand-over-hand locking (\autoref{ch:finegrained}).
\autoref{fig:queuefix} shows an implementation
that uses hand-over-hand locking
both for <{put}>() and for <{get}>().  It passes all tests.

@@

As in Hoare monitors, there is a hidden mutex associated with each Mesa monitor,
and the mutex must be acquired upon entry to a method and released
upon exit.
Mesa monitors also have condition variables that a thread can wait on.
Like in Hoare monitors, the \texttt{wait} operation releases the mutex.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call the corresponding Mesa
operation \texttt{notify} rather than \texttt{signal}.
\index{notify}%
When a thread <{p}> invokes \texttt{notify}, it does not immediately pass 
control to a thread that is waiting on the corresponding condition (if there
is such a thread).  Instead, <{p}> remains in the critical section
until it leaves the monitor explicitly
(by calling either \texttt{release} or \texttt{wait}).
At that point, any thread that
was notified will have a chance to enter the critical section, but they compete
with other threads trying to enter the critical section.

@@

Basically, there is just one gate to enter the critical section, instead
of a main gate and a gate per waiting condition.
This is a very important difference.  In Hoare monitors, when a thread
enters through a waiting gate, it can assume that the condition associated
with the waiting gate still holds because no other thread can run in between.
Not so with Mesa monitors: by the time a thread that was notified enters
through the main gate, other threads may have entered first and falsified
the condition.  So, in Mesa, threads always have to check the condition
again after resuming from the \texttt{wait} operation.  This is accomplished
by wrapping each \texttt{wait} operation in a \textbf{while} statement that
loops until the condition of interest becomes valid.
A Mesa monitor therefore is more closely related to busy waiting
than to split binary semaphores.

@@

Note that \texttt{wait} is always invoked within a \textbf{while}
loop that checks for the condition that the thread is waiting for.
It is \emph{imperative} that there is always a \textbf{while} loop
around any invocation of \texttt{wait} containing the negation of
the condition that the thread is waiting for.  Many implementation
of Mesa condition variables depend on this, and optimized implementations
of condition variables often allow so-called ``spurious wakeups,''
where \texttt{wait} may sometimes return even if the conditon
variable has not been notified.
As a rule of thumb, one should always be able to replace \texttt{wait}
by \texttt{unlock} followed by \texttt{lock}.
This turns the solution into a busy-waiting one, inefficient but still
correct.

@@

\texttt{write\_release} executes \texttt{notifyAll}(?\textit{r\_cond})
as well as \texttt{notify}(?\textit{w\_cond)}.  Because we
do not keep track of the number of waiting readers or writers, we
have to conservatively assume that all waiting readers can enter,
or, alternatively, up to one waiting writer can enter.  So
\texttt{write\_release} wakes up all potential candidates.  There
are two things to note here.  First, unlike split binary semaphores
or Hoare monitors, where multiple waiting readers would have to be
signaled one at a time in a baton-passing fashion (see
\autoref{fig:RWsplitsema}), with Mesa monitors all readers are
awakened in one fell swoop using \texttt{notifyAll}.  Second, both
readers and writers are awakened---this is ok because both execute
\texttt{wait} within a \textbf{while} loop, re-checking the condition
that they are waiting for.  So, if both type of threads are waiting,
either all the readers get to enter next or one of the writers gets
to enter next.  (If you want to prevent waking up both readers and
a writer, then you can keep track of how many threads are waiting
in the code.)

@@

To implement a \emph{No Hold and Wait} solution, a philosopher would need a
way to lock both the left and right forks at the same time.  Locks do not
have such an ability, and neither do semaphores. so we re-implement the
Dining Philosophers using condition variables that allow one to wait for
arbitrary application-specific conditions.
%
\autoref{fig:dinerscv} demonstrates how this might be done.
We use a single mutex for the diners, and, for each fork, a boolean
and a condition variable.  The boolean indicates if the fork has been
taken.
Each diner waits if either the left or right fork is already taken.
But which condition variable to wait on?
The code demonstrates an important technique to use when waiting for
multiple conditions.
\index{multiple conditions, waiting on}%
The condition in the \textbf{while} statement is the negation of the
condition that the diner is waiting and consists of two disjuncts.
Within the \textbf{while} statement,
there is an \textbf{if} statement for each disjunct.
The code waits for either or both forks if necessary.  After that, it goes
back to the top of the \textbf{while} loop.

@@

Try moving \texttt{lock}() to the beginning of the \texttt{increment} method
and \texttt{unlock}() to the end of \texttt{increment} and see what happens.
While Harmony will only report one faulty run, this incorrect code can lead to
the assertion failing as well as threads getting blocked indefinitely.

@@

\subsection*{\textbf{let}}

>>>>

\begin{problems}
\item Harmony programs can usually be easily translated into
Python by hand.  For example,
\autoref{fig:incpy} is a Python version of \autoref{fig:inc}.
\begin{enumerate}
\item Run \autoref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run \autoref{fig:incpy} 1000 times.
For example, if you are using the bash shell (in Linux or Mac OS X, say), you can do the following:
\begin{code}
\begin{verbatim}
for i in {1..1000}
do
    python Up.py
done
\end{verbatim}
\end{code}
If you're using Windows, the following batch script does the trick:
\begin{code}
\begin{verbatim}
FOR /L %%i IN (1, 1, 1000) DO python Up.py
PAUSE
\end{verbatim}
\end{code}
How many times does the assertion fail (if any)?
\end{enumerate}
\item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
incrementer thread increment <{count N}> times.  Run \autoref{fig:incmany}
10 times (using Python).
Report how many times the assertion fails and what the value of <{count}>
was for each of the failed runs.
Also experiment with lower values of <{N}>.
How large does <{N}> need to be for assertions to fail?
(Try powers of 10 for <{N}>.)
\item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
and run them through Harmony.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
(Also, you do not get to use the <{atomically}> keyword
or a \emph{lock}, yet.)
\end{problems}

@@

In Harmony, any statement can be preceded by the <{atomically}>
keyword.  It means that statement as a whole is to be executed atomically.
The <{atomically}> keyword can be used to specify the behavior of
methods such as <{acquire}> and <{release}>.  But
an actual executable program---such as the one in
\autoref{fig:cssynch}---should not use the <{atomically}> keyword
because---on a normal machine---it cannot be directly compiled into
machine code.
If we want to make the program executable on hardware,
we have to show how \texttt{Lock}, <{acquire}>, and <{release}>
are implemented, not just how they are specified.
\autoref{ch:spinlock} presents such an implementation.

@@

To see that invariant $\mathcal{I}_1$ is maintained, note that
\texttt{!}<{p}> = <{True}> upon entry of \texttt{test\_and\_set}
(because of the condition on the <{while}> loop that the
\texttt{test\_and\_set} method is invoked in).
There are two cases:
\begin{enumerate}
\item \texttt{!}<{s}> is <{False}> upon entry to \texttt{test\_and\_set}.
Then upon exit \texttt{!}<{p}> = <{False}> and \texttt{!}<{s}> = <{True}>,
maintaining the invariant.
\item \texttt{!}<{s}> is <{True}> upon entry to \texttt{test\_and\_set}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section
because we can assume, by the induction hypothesis, that $\textit{private}[i]$ is
<{True}> just before exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread <{i}> only proceeds to the critical
section if \textit{private}[<{i}>] is <{False}>, and (ii) no other thread modifies
\textit{private}[<{i}>].

@@

\autoref{fig:suspension} shows the lock interface using suspension.
It is implemented as follows:
\begin{itemize}
\item A lock maintains both a boolean indicating whether the
lock is currently acquired and a list of contexts of threads that want to
acquire the lock.
\item
<{acquire}>()
\index{acquire}%
acquires the lock if available and suspends the invoking thread if not.
In the latter case, the context of the thread is added to the end of the list of contexts.
Note that \textbf{stop} is called within an atomic statement block---this is
the only exception to such an atomic statement block running to completion.
While the
thread is running no other threads can run, but when the thread suspends itself
other threads can run.
\item
<{release}>()
\index{release}%
checks to see if any threads are waiting to acquire the lock.
If so, it uses the <{head}> and <{tail}>
methods from the \texttt{list} module (see \autoref{ap:module})
to resume the first thread that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first thread is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
<{choose}> to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents starvation.

@@

\section*{Exercises}
\begin{problems}
\item
Run \autoref{fig:incfixed} using (i) <{synch}> and (ii) \texttt{synchS}.
Report how many states were explored by Harmony for each module.
\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables <{x}>
(initially 0) and <{y}> (initially 100) that can be accessed through methods
\texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
should return a pair that sums to 100.  Add the necessary synchronization code.
\item \label{ex:trylock} Implement \texttt{tryAcquire}(\textit{b}) as an additional
interface for both the <{synch}> and \texttt{synchS} modules.
This interface is like <{acquire}>(\textit{b}) but never blocks.  It
returns <{True}> if the lock was available (and now acquired) or <{False}>
if the lock was already acquired.
Hint: you do not have to change the existing code.
\item People who use an ATM often first check their balance and then withdraw
a certain amount of money not exceeding their balance.  A negative balance
is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
one to check the balance and one to withdraw money.
Note that all operations on accounts are carefully protected by a lock
(i.e., there are no data races).
The \texttt{customer}
method models going to a particular ATM and withdrawing money not exceeding
the balance.
Running the code through Harmony reveals that there is a bug.
It is a common type of concurrency bug known as \emph{Time Of Check Time Of
Execution} (TOCTOE).
\index{Time Of Check Time Of Execution}
\index{TOCTOE}
In this case, by the time the withdraw operation is performed,
the balance can have changed.
Fix the code in \autoref{fig:atm}.  Note, you should leave the
customer code the same. You are only allowed to change the
\texttt{atm\_} methods, and you cannot use the \textbf{atomically}
keyword.
\end{problems}

@@

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queueconc}
uses the <{alloc}> module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using \texttt{malloc}() and \texttt{free}().
\texttt{malloc}(<{v}>) returns a new memory location initialized to <{v}>,
which should be released with \texttt{free}() when it is no longer in use.
The queue maintains a <{head}> pointer to the first element in the list
and a <{tail}> pointer to the last element in the list.
The <{head}> pointer is <{None}> if and only if the queue is empty.
(<{None}> is a special address value that is not the address of any
memory location.)

@@

\autoref{fig:qtest3} checks a mixture of
concurrent <{get}> and <{put}> operations.
It starts <{N getter}> threads and
<{N}> \texttt{putter} threads.
Thread <{main}> waits until all have finished,
and then checks to make sure that whatever is left
on the queue was not also consumed by some
<{getter}> thread.
Finally, it makes sure that the queue is empty
once all the values that were added
to it by the \texttt{putter} threads have been consumed.

A common mistake that people make is to use
a globally shared variable in the implementation of
a concurrent data structure.
The <{get()}> method of the queue implementation
uses a local variable <{node}>, declared using
a <{let}> statement, but it would be easy to
mistakenly use a global variable <{node}> instead.
If there is only one queue, this is not an issue.
So, it is important to test not just one queue at a
time, but also multiple queues.
\autoref{fig:qtest4} gives an example that
tests the <{get}> method for use of global
variables.  Even if the assertion does not fail,
Harmony might find a data race if there are global
variables shared between different queues.

@@

So, you wrote a Harmony program and Harmony reports a problem.
Often you may just be able to figure it out by staring at the code and
going through some easy scenarios, but what if you don't?
The output of Harmony can be helpful in that case.

@@

Let us run the code through the test programs in the last
chapter.  Harmony does not detect any issues with the sequential test
in \autoref{fig:qtestseq}.
(Run this using the \texttt{-m} flag like this:
\texttt{harmony -m queue=queuebroken code/qtestseq.hny})
However, when we run
the new queue code through the test in \autoref{fig:qtestpar}, Harmony
reports a safety violation (even without specifying a behavior).
The command line to reproduce this is:

\begin{code}
\begin{verbatim}
harmony -m queue=queuebroken code/qtestpar.hny
\end{verbatim}
\end{code}

Before we go look at the details of what went wrong, we want to make
sure that we generate the simplest scenario.  So, first we want to explore
what the smallest <{NOPS}> (number of operations or number of
threads) that causes the bug to occur.  With some experimentation,
we find that <{NOPS = 2}> does not find a problem, but
<{NOPS = 3}> does
(\texttt{harmony -m queue=queuebroken -c NOPS=3 code/qtestpar.hny})).
\autoref{fig:queuebroken1} shows the Harmony output.

There is quite a bit of information in the Harmony output, and while
it may seem intimidating, we have to learn to navigate through it
step-by-step.
Let's start with looking at the red text.  Harmony found a safety
violation (something bad happened during one of the possible executions),
and in particular <{thread}>(2) (thread T2) was trying to dereference
the address <{?alloc["pool"][0]["lock"]}> (aka <{?alloc.pool[0].lock}>).

The <{alloc}> module maintains a shared array <{pool}> that
it uses for dynamic allocation.  Apparently T2 tried to access
<{pool}>[0], but it does not exist, meaning that either it was not
yet allocated, or it had been freed since it was allocated.
When we look at the top half of the figure, we see that in fact thread
T1 allocated <{pool}>[0] in turn 2, but T3 freed it in turn 4.
Looking back down, we see that T1 executed <{thread(1)}> and has
since terminated, while T3 is executing <{thread(3)}>.

Looking further at the stack traces, we can see that
T3 was in the process of executing
<{release(?q.lock)}> within <{get(?q}>).
T1 is currently executing <{acquire(?alloc.pool[0].lock)}>
within <{put(?q, 2)}>, but 
<{alloc.pool}>[0] does not exist.
The corresponding line of Harmony code is
\textbf{atomically when not} !\textit{binsema}
in line 25 of the \texttt{sync} module.

So, how did we get there?  In the top we can see that the order of
events was the following:
\begin{enumerate}
\item[0:] initialization completed, with <{q}> being
<{{ .lock: False, .next: None, .value: None }}>;
\item[1:] thread T1 (<{thread(1)}>) ran and finished executing
<{put(1)}> (see the output column for that clue: the thread printed that).
We can see that <{q.next}> now points to <{alloc.pool[0]}>,
which the thread must have allocated.  The contents is 
<{{ .lock: False, .next: None, .value: 1 }}>, as expected;
\item[2:] thread T2 (<{thread(1)}>) started running, calling
<{put(?q, 2)}>.  We can see it got as far as putting 2 on the queue,
but it is not yet done.  It is currently trying to acquire
<{alloc.pool[0].lock}>;
\item[3:] thread T3 (<{thread(1)}>) started running, calling
<{get(?q)}>.  We can also see that it freed <{pool[0]}>,
and is now releasing <{q.lock}>;
\item[4:] thread T2 resumes and tries to access <{pool[0]}>, which
no longer exists (because T3 just freed it).
\end{enumerate}

Clearly there was a race in which T2 was trying to lock
<{pool[0].lock}> (which contained the node with the value 1)
while T3 was freeing that very same node, and T2 lost the race.
More precisely, T2 was executing <{put(?q, 2)}>, when T3
preempted it with <{get(?q)}> and removed the node that T2
was trying to access.  But why did the locks not prevent this?

\index{reserve debugging}%
It is time to start stepping through the code that has been executed
before this happened.  This is sometimes known as
\emph{reverse debugging}.  In fact, Harmony allows you to step through
an execution forwards and backwards.
In this case, we first want to see what
T2 is doing.  You can click on its first (top-left) orange box to
time-travel to that part in the execution.
Now by hitting $\langle\mathtt{return}\rangle$ repeatedly,
we can quickly skip through the code.
T2 first calls <{put(?q, 1)}> and then allocates a new
node initialized with a lock.  Keep stepping until it executes <{nq = q}>.
Hit $\langle\mathtt{return}\rangle$ once more and inspect the
state of T2 in the lower-right corner.  You can see that variable
<{nq}> is initialized to~<{?q}>.  T2 then enters into the
<{while}> loop and tries to acquire <{nq->lock}>.
This succeeds, and next T2 executes <{let n = nq->next}>.
Now <{n = ?alloc.pool[0]}>, which is not <{None}>.
It then releases <{nq->lock}>
(<{nq}> points to <{q}>).  It then sets <{nq}> to $n$, which
is still <{alloc.pool[0]}>.  Finally, it calls <{acquire(?nq->lock)}>.
But before it can complete that operation, T3 runs next.

T3 chooses <{"get"}> and then goes on to invoke <{get(?q)}>.
This first successfully acquires <{q}->lock}>.
T3 then finds out that <{q->next}> points to <{alloc.pool[0]}>.
T3 sets <{node}> to <{alloc.pool[0]}> as well and sets
<{q->next}> to <{node->next}>.
T3 sets <{result}> to <{node->value}> (which is 1)
and then frees <{node}>.  This is where the problem is---T2 is
about to acquire the lock in that same node.

To fix the code without changing the data structure, we can use
hand-over-hand locking (\autoref{ch:finegrained}).
\autoref{fig:queuefix} shows an implementation
that uses hand-over-hand locking
both for <{put()}> and for <{get()}>.  It passes all tests.

@@

As in Hoare monitors, there is a hidden mutex associated with each Mesa monitor,
and the mutex must be acquired upon entry to a method and released
upon exit.
Mesa monitors also have condition variables that a thread can wait on.
Like in Hoare monitors, the \texttt{wait} operation releases the mutex.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call the corresponding Mesa
operation \texttt{notify} rather than \texttt{signal}.
\index{notify}%
When a thread <{p}> invokes \texttt{notify}, it does not immediately pass 
control to a thread that is waiting on the corresponding condition (if there
is such a thread).  Instead, <{p}> remains in the critical section
until it leaves the monitor explicitly
(by calling either <{release}> or \texttt{wait}).
At that point, any thread that
was notified will have a chance to enter the critical section, but they compete
with other threads trying to enter the critical section.

@@

Basically, there is just one gate to enter the critical section, instead
of a main gate and a gate per waiting condition.
This is a very important difference.  In Hoare monitors, when a thread
enters through a waiting gate, it can assume that the condition associated
with the waiting gate still holds because no other thread can run in between.
Not so with Mesa monitors: by the time a thread that was notified enters
through the main gate, other threads may have entered first and falsified
the condition.  So, in Mesa, threads always have to check the condition
again after resuming from the \texttt{wait} operation.  This is accomplished
by wrapping each \texttt{wait} operation in a <{while}> statement that
loops until the condition of interest becomes valid.
A Mesa monitor therefore is more closely related to busy waiting
than to split binary semaphores.

@@

Note that \texttt{wait} is always invoked within a <{while}>
loop that checks for the condition that the thread is waiting for.
It is \emph{imperative} that there is always a <{while}> loop
around any invocation of \texttt{wait} containing the negation of
the condition that the thread is waiting for.  Many implementation
of Mesa condition variables depend on this, and optimized implementations
of condition variables often allow so-called ``spurious wakeups,''
where \texttt{wait} may sometimes return even if the conditon
variable has not been notified.
As a rule of thumb, one should always be able to replace \texttt{wait}
by <{unlock}> followed by <{lock}>.
This turns the solution into a busy-waiting one, inefficient but still
correct.

@@

\texttt{write\_release} executes \texttt{notifyAll}(?\textit{r\_cond})
as well as \texttt{notify}(?\textit{w\_cond)}.  Because we
do not keep track of the number of waiting readers or writers, we
have to conservatively assume that all waiting readers can enter,
or, alternatively, up to one waiting writer can enter.  So
\texttt{write\_release} wakes up all potential candidates.  There
are two things to note here.  First, unlike split binary semaphores
or Hoare monitors, where multiple waiting readers would have to be
signaled one at a time in a baton-passing fashion (see
\autoref{fig:RWsplitsema}), with Mesa monitors all readers are
awakened in one fell swoop using \texttt{notifyAll}.  Second, both
readers and writers are awakened---this is ok because both execute
\texttt{wait} within a <{while}> loop, re-checking the condition
that they are waiting for.  So, if both type of threads are waiting,
either all the readers get to enter next or one of the writers gets
to enter next.  (If you want to prevent waking up both readers and
a writer, then you can keep track of how many threads are waiting
in the code.)

@@

To implement a \emph{No Hold and Wait} solution, a philosopher would need a
way to lock both the left and right forks at the same time.  Locks do not
have such an ability, and neither do semaphores. so we re-implement the
Dining Philosophers using condition variables that allow one to wait for
arbitrary application-specific conditions.
%
\autoref{fig:dinerscv} demonstrates how this might be done.
We use a single mutex for the diners, and, for each fork, a boolean
and a condition variable.  The boolean indicates if the fork has been
taken.
Each diner waits if either the left or right fork is already taken.
But which condition variable to wait on?
The code demonstrates an important technique to use when waiting for
multiple conditions.
\index{multiple conditions, waiting on}%
The condition in the <{while}> statement is the negation of the
condition that the diner is waiting and consists of two disjuncts.
Within the <{while}> statement,
there is an \textbf{if} statement for each disjunct.
The code waits for either or both forks if necessary.  After that, it goes
back to the top of the <{while}> loop.

@@

Try moving <{lock()}> to the beginning of the <{increment}> method
and <{unlock()}> to the end of <{increment}> and see what happens.
While Harmony will only report one faulty run, this incorrect code can lead to
the assertion failing as well as threads getting blocked indefinitely.

@@

\subsection*{<{let}>}
